---
title: "Simple multinomial logistic regression using BDA"
author: Joseph V. Casillas
date: 2024-02-24
format:
  html:
    theme: united
---

First, let's load some libraries and setup our document. 

```{r}
#| label: setup
#| message: false

# Doc setup
knitr::opts_chunk$set(
  echo = TRUE, 
  fig.asp = 9/16, 
  fig.algn = "center", 
  out.width = "100%"
)

# Load libs
library("tidyverse")
library("ggdist")
library("brms")
library("here")
library("scales")
library("marginaleffects")
library("tinytable")
library("glue")

# Set seed for reproducibility
set.seed(20240224)

# Set theme for plots
theme_set(theme_light(base_family = "Palatino"))

# Set some colors
cols <- c( "#A7C7E7" = "blue", "#355E3B" = "green", "#cc0033" = "red")
```

Next, we simulate some data for the analysis. 

```{r}
#| label: data-sim

# dataframe variables
n_ids       <- 50
n_lang      <- 2
n_item_reps <- 15
probs_en    <- c(0.2, 0.5, 0.8)
probs_sp    <- c(0.8, 0.2, 0.5)
choices     <- c("Green", "Blue", "Red")

# Set up dataframe
dat <- tibble(
  id = rep(glue::glue("id_{1:n_ids}"), times = n_lang), 
  Language = rep(c("English", "Spanish"), each = n_ids), 
  Response = c(
    sample(x = choices, size = n_ids, replace = TRUE, prob = probs_en), 
    sample(x = choices, size = n_ids, replace = TRUE, prob = probs_sp)
  )
)

```

We are going to pretend we did some sort of experiment in which we asked bilingual people their favorite cup out of three options: a red cup, a green cup, and a blue cup. 
We will assume the cups are exactly the same except for the colors. 
A priori, there is no reason to think that people would have different preferences based on which language they are asked in, but this is made up data, so we'll ignore that for now. 
Let's take a look at what we made.

```{r}
#| label: get-to-know-data

head(dat)
glimpse(dat)
```

We have a tidy data frame with 3 columns (i.e., 3 variables). 
There is data from 50 individuals, in each of their languages (Spanish, English), leaving us with 50 $\times$ 2 = 100 observations. 
Now let's get some general descriptives.

```{r}
#| label: tbl-simple-descriptives
#| tbl-cap: Summary of responses

dat |> 
  count(Response) |> 
  mutate(Total = sum(n), Prop = n / Total) |>
  tt(theme = "bootstrap", width = 0.7) |> 
  style_tt(j = 2:4, align = "r")
```

It's clear we see a higher propensity to select `red` overall, but let's make a plot for fun. 

```{r}
#| label: fig-simple
#| fig-cap: Probability of a given response in the data set.

dat |> 
  count(Response) |> 
  mutate(
    total = sum(n), 
    prop = n / total, 
    Response = fct_reorder(Response, prop)
  ) |> 
  ggplot() + 
  aes(x = prop, color = Response) + 
  geom_segment(
    aes(x = 0, xend = prop, y = Response, yend = Response), 
    lineend = "round", linewidth = 17, show.legend = F, alpha = 0.5
    ) + 
  geom_segment(
    aes(x = 0, xend = prop, y = Response, yend = Response), 
    lineend = "round", linewidth = 13, show.legend = F
    ) + 
  geom_text(aes(y = Response, label = prop), color = "white", show.legend = F) + 
  scale_color_manual(values = names(cols)) + 
  scale_x_continuous(breaks = seq(0, 1, 0.1), expand = c(0, 0, 0, 0.05)) + 
  coord_cartesian(xlim = c(0, 1)) + 
  labs(y = NULL, x = "Proportion of responses") 

```

Ok. 
Looks good. 
We see that, overall, `red` was selected most, followed by `green`, and then `blue`. 
Now we can build our first Bayesian model.
We will fit a simple, intercept-only model, ignoring `language` for now. 
The specification will be as follows: 

$$
\begin{aligned}
&\ \textbf{Multinomial probability of selection of color}_i \\
\text{Color}_{i} \sim&\ \operatorname{Categorical}(\{\mu_{blue,i}, \mu_{green,i}, \mu_{red,i}\}) \\[10pt]
&\ \textbf{Model for probability of each option} \\
\{\mu_{blue,i}, \mu_{green,i}, \mu_{red,i}\} =&\ \beta_0 \\[5pt]
&\ \textbf{Priors} \\
\beta_{0} \sim&\ \mathcal{N} (0, 2) \qquad\qquad\ \ \text{Prior for choice-level coefficients} \\
\end{aligned}
$$

Essentially, the goal of this first model is to get posterior estimates of the probability of selecting `red`, `green`, or `blue`, which should align with what we just saw in @tbl-simple-descriptives and @fig-simple.
We specify the model as follows.

```{r}
#| label: mod0

# This is the BRMS syntax of the model highlighted above
form <- bf(Response ~ 0 + Intercept)

# Use get_prior if you need help figuring out the priors
# get_prior(formula = form, family = categorical(link = "logit"), data = dat)
#  prior class      coef     dpar        source
# (flat)     b            muGreen       default
# (flat)     b Intercept  muGreen  (vectorized)
# (flat)     b              muRed       default
# (flat)     b Intercept    muRed  (vectorized)


# Set priors
simple_priors <- c(
  prior(normal(0, 2), class = b, dpar = muGreen), 
  prior(normal(0, 2), class = b, dpar = muRed)
)

# Fit simple model
b_multi_0 <- brm(
  formula = form, 
  prior = simple_priors, 
  iter = 2000, warmup = 1000, cores = 4, 
  data = dat, 
  family = categorical(link = "logit"), 
  file = here("b_multi_0")
)

# Get posterior and store in vector
mod0_post <- predictions(model = b_multi_0) |>
  posterior_draws() 
```

We can take a look at the model output using `summary()`, but it often isn't very helpful with this type of model. 

```{r}
#| label: print-simple-mod
summary(b_multi_0)
```

Let's use summarize the posterior to get an idea of probability of each choice. 
For fun we will make a table and a plot just like @tbl-simple-descriptives and @fig-simple. 
They should be essentially the same, but now we are summarizing an entire distribution of estimates, so we can include uncertainty in @tbl-mod0 and @fig-mod0.

```{r}
#| label: tbl-mod0
#| tbl-cap: Posterior summary of intercept-only model

mod0_post |> 
  group_by(group) |> 
  median_qi(x = draw) |> 
  transmute(
    Response = group, 
    Estimate = glue("{round(x, 3)} [{round(.lower, 3)}, {round(.upper, 3)}]")
  ) |> 
  tt(theme = "bootstrap", width = 0.7,  
     notes = "Posterior median and 95% CI") |> 
  style_tt(j = 2, align = "r")

```


```{r}
#| label: fig-mod0
#| fig-cap: Posterior medians of color choice along with 95% and 66% credible intervals. 

mod0_post |> 
  ggplot() + 
  aes(x = draw, y = group, fill = group) + 
  stat_slab(show.legend = F) + 
  stat_pointinterval(
    show.legend = F, pch = 21, position = position_nudge(y = -0.05)
    ) + 
  scale_fill_manual(values = names(cols)) + 
  scale_x_continuous(breaks = seq(0, 1, 0.1), labels = label_percent()) + 
  coord_cartesian(xlim = c(0, 1)) + 
  labs(y = NULL, x = "p(Color)")

```

It worked!
Looks good. 
Now we will fit a more complex model in which we include the language in which the person was asked, Spanish or English, but first let's look at the descriptives. 

```{r}
#| label: tbl-complete-descriptives
#| tbl-cap: Summary of responses as a function of language (Spanish, English)

dat |> 
  group_by(Language) |> 
  count(Response) %>% 
  mutate(
    total = sum(n), prop = n / total, 
    Language = case_when(
      Language == "English" & Response == "Blue" ~ "English", 
      Language == "Spanish" & Response == "Blue" ~ "Spanish", 
      TRUE ~ ""
    )
  ) |> 
  tt(theme = "bootstrap", width = 0.7) |> 
  style_tt(j = 2:4, align = "r")
```

```{r}
#| label: fig-complete
#| fig-cap: Average choice (±SE) as a function of language (Spanish, English)

dat |> 
  group_by(Language) |>
  count(Response) |> 
  mutate(
    total = sum(n), 
    prop = n / total, 
    se = sd(prop) / sqrt(n), 
    .groups = "drop"
    ) |> 
  ggplot() + 
  aes(y = Response, x = prop, color = Response, shape = Language) + 
  geom_linerange(
    aes(xmin = prop - se, xmax = prop + se, y = Response),
    linewidth = 4, lineend = "round", alpha = 0.5, 
    position = position_dodge(0.5), show.legend = T
    ) + 
  geom_point(size = 3, stroke = 2.5, position = position_dodge(0.5)) +
  scale_color_manual(name = NULL, values = names(cols), guide = "none") + 
  scale_shape_manual(name = NULL, values = 21:22) + 
  scale_x_continuous(breaks = seq(0, 1, 0.1), labels = label_percent()) + 
  coord_cartesian(xlim = c(0, 1)) + 
  labs(y = NULL, x = NULL) + 
  theme(legend.position = c(1, 0), legend.justification = c(1, 0), 
        legend.background = element_blank())

```

Great. 
So @tbl-complete-descriptives and @fig-complete give us an idea of what's going on (big differences in responding "red" or "green" based on language... again, make-up data). 
Now we will fit a model that accounts for language. 
Because language is a within participants variable (every participant was asked in English and Spanish) we will including a grouping variable for `id` with a varying slope of `language`. 
The model formulation is as follows: 

$$
\begin{aligned}
&\ \textbf{Multinomial probability of selection of color}_i \textbf{ in respondent}_j \\
\text{Choice}_{i_j} \sim&\ \operatorname{Categorical}(\{\mu_{1,i_j}, \mu_{2,i_j}, \mu_{3,i_j}\}) \\[10pt]
&\ \textbf{Model for probability of each option} \\
\{\mu_{1,i_j}, \mu_{2,i_j}, \mu_{3,i_j}\} =&\ (\beta_0 + b_{0_j}) + \beta_1 \text{Language[Spanish]}_{i_j} \\[5pt]
b_{0_j} \sim&\ \mathcal{N}(0, \sigma_0) \qquad\quad\quad \text{Respondent-specific offsets from global probability} \\[10pt]
&\ \textbf{Priors} \\
\beta_{0 \dots 1} \sim&\ \mathcal{N} (0, 2) \qquad\qquad\ \ \text{Prior for choice-level coefficients} \\
\sigma_0 \sim&\ \operatorname{Exponential}(1) \quad \text{Prior for between-respondent variability}
\end{aligned}
$$

Now we fit the model in `brms`. 

```{r}
#| label: mod1

# Specify model formula
form2 <- bf(Response ~ 0 + Intercept + Language + (1 + Language | ID | id))

# Use get_prior with unsure about priors
# get_prior(formula = form2, family = categorical(link = "logit"), data = dat) 
# class            coef         dpar                 source
#   cor                                             default
#   cor                                         (vectorized)
#     b                      muGreen                default
#     b       Intercept      muGreen            (vectorized)
#     b LanguageSpanish      muGreen            (vectorized)
#    sd                      muGreen                default
#    sd                      muGreen            (vectorized)
#    sd       Intercept      muGreen            (vectorized)
#    sd LanguageSpanish      muGreen            (vectorized)
#     b                        muRed                default
#     b       Intercept        muRed            (vectorized)
#     b LanguageSpanish        muRed            (vectorized)
#    sd                        muRed                default
#    sd                        muRed            (vectorized)
#    sd       Intercept        muRed            (vectorized)
#    sd LanguageSpanish        muRed            (vectorized)

# Fit model
b_multi_1 <- brm(
  formula = form2,  
  prior = c(
    prior(normal(0, 2), class = b, dpar = muGreen), 
    prior(normal(0, 2), class = b, dpar = muRed), 
    prior(exponential(1), class = sd, dpar = muGreen), 
    prior(exponential(1), class = sd, dpar = muRed), 
    prior(lkj(1), class = cor)
  ), 
  iter = 2000, warmup = 1000, cores = 4, 
  data = dat, 
  family = categorical(link = "logit"), 
  file = here("b_multi_1")
  )

```

Let's take a look at the model summary: 

```{r}
#| label: print-mod1
summary(b_multi_1)
```

And a plot to see how our model compares to the descriptives: 

```{r}
#| label: fig-mod1
#| fig-cap: Posterior medians of color choice as a function of language along with 95% and 66% credible intervals

avg_predictions(
  model = b_multi_1, 
  variables = "Language", 
  re_formula = NULL
  ) |>
  posterior_draws() |> 
  ggplot() + 
  aes(x = draw, y = group, fill = group, group = Language, shape = Language) + 
  stat_halfeye(position = position_dodge(0.5), point_fill = "white") + 
  scale_fill_manual(name = NULL, values = names(cols), guide = "none") + 
  scale_shape_manual(name = NULL, values = 21:22) + 
  scale_x_continuous(breaks = seq(0, 1, 0.1), labels = label_percent()) + 
  coord_cartesian(xlim = c(0, 1)) + 
  labs(y = NULL, x = "p(Color|Language)") + 
  theme(legend.position = c(1, 0), legend.justification = c(1, 0), 
        legend.background = element_blank())

```

It looks like @fig-mod1 captures what we saw qualitatively in @tbl-complete-descriptives and @fig-complete. 
Let's finish by comparing the difference between languages for each color. 

```{r}
#| label: fig-diff-plot
#| fig-cap: Marginal mean difference of color preference (Spanish $-$ English).

avg_slopes(
    model = b_multi_1, 
    variables = "Language", 
    re_formula = NA
  ) |> 
  posterior_draws() |> 
  ggplot() + 
  aes(x = draw, y = group, fill = group) + 
  geom_vline(xintercept = 0, lty = 3) + 
  stat_halfeye(show.legend = F, pch = 21, point_fill = "white") + 
  scale_fill_manual(name = NULL, values = names(cols)) + 
  scale_x_continuous(breaks = seq(-1, 1, 0.2), labels = label_percent()) + 
  coord_cartesian(xlim = c(-1, 1)) + 
  labs(y = NULL, x = "Marginal mean difference:\nSpanish - English")
```

